Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark

### 2021-02-03 reports/article.Rmd
updated article with only 90 days

# Results -------------------------------------------------------------
crude_rate_90d <- sum(df$outcome_infection_90d) / nrow(df)
crude_rate_2y  <- sum(df$outcome_infection_2y) / nrow(df)

get_auc <- function(Time, model, version = "1") {
  data_auc_ci %>% 
  filter(Period == Time, Model == model) %>% 
  select(ends_with(version)) %>% 
  pluck(1)
}
```

@@

**PARTICIPANTS AND METHODS:** We analysed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden. Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days or two years after surgery. Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from **XXX**. We assessed model performance by analysing the area under the receiver operation characteristics curve (AUC) and by calibration belt plots.

**RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate_90d * 100` % within 90, days and `r crude_rate_2y * 100` % within two years in the Swedish derivation cohort. A prediction model for PJI within 90 days combined `r coefs_text["90 days"]`, resulting in an `r get_auc("90 days", "Reduced model", "2")` internally and **(AUC=XXX, 95% CI: XXX to XXX)** externally. A model for PJI within two years had similar performance internally `r get_auc("2 years", "Reduced model")` and externally **XXX (95% CI: XXX to XXX)**. These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("90 days", "ASA class")`, Charlson `r get_auc("90 days", "Charlson")`, Elixhauser `r get_auc("90 days", "Elixhauser")`, and the Rx Risk V `r get_auc("90 days", "Rx Risk V")` comorbidity indices.

@@

We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration. The best models were then validated externally on a cohort from the **XXX**.

@@

## External validation cohort (XXX)

@@

The Charlson, Elixhauser and Rx Risk V comorbidity indices are based on the presence of ICD-10 and ATC codes grouped into main categories (`r nrow(coder::charlson)`, `r nrow(coder::elixhauser)`, and `r nrow(coder::rxriskv)` categories, respectively).[@Quan2005; @Pratt2018] Diagnostic categories from all indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab. \@ref(tab:tabcategorization)). Some of the combined comorbidities were observed for less than 10 patients with or without PJI, and these diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events. PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-code recorded in the NPR within 90 days or 2 years (Tab. \@ref(tab:infcodes)), or if the patient was reported to SHAR as re-operated due to infection within the same time frame.

@@

For each outcome (PJI within 90 days or 2 years), ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in a logistic LASSO regression model. We kept the $\lambda_i$ that minimized the mean cross-validated deviance. 100 bootstrap samples were then drawn from the observed data set.[@Austin2004]

Logistic LASSO regression was performed for each sample using the chosen $\lambda_i$ as penalty term. Regression coefficients were ranked based on their absolute values as a measure of variable importance. We then used Bayesian methods to estimate a posterior distribution and to identify two change points in a linear segmented piecewise regression model. The second change point was then used as a break-point. Influential predictors with absolute coefficients value above this point, scored a point. The whole process, each with 100 Bootstrap replicates, was repeated 100 times (yielding 10,000 bootstrap samples in total). Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables. Even more parsimonious models with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison. Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison. Odds ratios for the final models were estimated with 95 % confidence intervals.

@@

Each developed model was used to predict the probability of PJI within 90 days or two years for patients derived from the SHAR (internal validation). Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples. The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed. Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a **XXX** cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to **XXX**. Calibration for this re-calibrated model was illustrated in the same calibration belt plot as after internal calibration.

@@

We built an online web calculator available at https **XXX** to be used in clinical practice. R software (version 3.6.1; R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.

@@

We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). `r crude_rate_90d * 100` % (n = `r sum(df$outcome_infection_90d)`) of the patients developed a PJI within 90 days, and `r crude_rate_2y * 100` % (n = `r sum(df$outcome_infection_2y)`) within 2 years (Tab. \@ref(tab:tab1)). The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V. In addition, **XXX** patients from **XXX** were included as an external validation cohort (Fig. \@ref(fig:flowchart), right panel).

@@

Patients with `r tolower(excl_factors)` were uncommon, wherefore those comorbidities were excluded as potential predictors prior to further modeling. ROC-curves (Fig. \@ref(fig:rocs)) and AUC-values (Fig. \@ref(fig:aucci)) for the main and reduced models were very similar, and for simplicity, we therefore focused on the reduced models. Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilities for this outcome (Fig. \@ref(fig:sep)).

The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text["90 days"]` (Tab. \@ref(tab:coefs)). This reduced model for the predicition of PJI within 90 days had an `r get_auc("90 days", "Reduced model", "2")`, whereas ASA class `r get_auc("90 days", "ASA class")`, the Charlson comorbidity index `r get_auc("90 days", "Charlson")`, the Elixhauser comorbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` had less discriminative ability (Fig. \@ref(fig:rocs) and \@ref(fig:aucci)). The reduced model had good accuracy as visualized by the calibration plot (Fig. \@ref(fig:calibration), left panel).

The reduced model for PJI within two years included the same predictors, with the addition of `r coefs_text["2 years"]` (Tab. \@ref(tab:coefs)). Predictors conferring a statistically significant risk increase were those **mentioned as predictive for PJI within 90 days, excluding osteoporosis or Paget's disease, and additionally included an underlying diagnosis of AVN or inflammatory joint disease, and the presence of CNS disease**. The reduced model for the prediction of PJI within two years had an `r get_auc("2 years", "Reduced model", "2")`, whereas ASA class `r get_auc("2 years", "ASA class")`, the Charlson comorbidity index `r get_auc("2 years", "Charlson")`, the Elixhauser comorbidity index `r get_auc("90 days", "Elixhauser")`, and Rx Risk V `r get_auc("90 days", "Rx Risk V")` again had less discriminative ability (Fig. \@ref(fig:rocs) and \@ref(fig:aucci)). Calibration was good for the reduced model (Fig. \@ref(fig:calibration), right panel).

@@

We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices. A similarly developed model to predict PJI within two years included similar predictors as the model developed to predict PJI within 90 days, but added `r coefs_text["2 years"]` to the list of predictor variables. This model also performed better than the established comorbidity measures in terms of AUC and calibration.

@@

Since the prediction models both for PJI within 90 days and within two years were based on easily accessible information they can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab. \@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics. For example, `r examples$desc[[1]]` for PJI within 90 days; `r examples$desc[[2]]`.

@@

The predictors in our models were chosen based on a strict analytical approach, not on proposed relevance of patient or procedure characteristics. The majority of the identified predictor variables are however associated with the risk of developing PJI in previously developed prediction models and within the setting of observational studies, which is supportive of our own models.

@@

NPH initiated the study, OR managed the ethical review board application in Sweden, EB developed the statistical model, **XXX** performed external validation with data from **XXX**. NPH and EB drafted, and all authors edited and finalized the manuscript.

# Competing interest statement

All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi\_disclosure.pdf and declare ...

# Transparency Statement

NH, the principal investigator of this study, affirms that this is an accurate report with no relevant data or aspects omitted.

# Public and patient involvement statement

There is a patient representative at SHAR to ensure the patients' interests are respected in planned research.

# Data sharing statement

Data is available upon reasonable request to the principal investigator.

@@

```{r brlprop}
load("../cache/infection_data.RData")

infection_data %>% 
  select(outcome, time, coefs_selected) %>% 
  mutate(coefs_selected = map(coefs_selected, "brlasso_tbl_selected")) %>% 
  unnest("coefs_selected") %>%
  filter(outcome == "infection") %>% 
  select(-outcome) %>% 
  pivot_wider(names_from = "time", values_from = "n") %>% 
  mutate(
    variable = clean_names(variable) %>% 
      {gsub("\\.", " ", .)} %>% 
      {gsub("Rheumatidisease", "Rheumatic disease", .)} %>% 
      {gsub("Sex Male", "Male gender", .)}
  ) %>% 
  select(Predictor = variable, `90 days` = `90d`, `2 years` = `2y`) %>% 
knitr::kable(
  caption = "Variables selected by the bootstrap ranking procedure, and therefore kept in the main model. Variables chosen at least 33 times out of 100 were kept in the reduced model as well. (BMI = body mass index. ASA class = American Society for Anesthesiologists classification. CNS = central nervous system.)")
```

@@

```{r coefs}
tbl_coefs_present %>% 
  mutate(
    time = replace(as.character(time), duplicated(time), ""),
    beta = round(beta, 2)
  ) %>% 
knitr::kable(
  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models. BMI = body mass index ("under/normal weight as baseline). CNS = central nervous system. Primary osteoarthritis was baseline for diagnosis and female was baseline for sex.'
)
```

>>>>

# Results -------------------------------------------------------------
crude_rate <- sum(df$outcome) / nrow(df)

get_auc <- function(model, version = "1") {
  data_auc_ci %>% 
  filter(Model == model) %>% 
  select(ends_with(version)) %>% 
  pluck(1)
}
```

@@

**PARTICIPANTS AND METHODS:** We analysed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden. Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days after surgery. Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from **XXX**. We assessed model performance by analysing the area under the receiver operation characteristics curve (AUC) and by calibration belt plots.

**RESULTS:** The crude incidence rate of PJI after THA was `r crude_rate * 100` % within 90 days in the Swedish derivation cohort. A prediction model for PJI within 90 days combined `r coefs_text`, resulting in an `r get_auc("Reduced model", "2")` internally and **(AUC=XXX, 95% CI: XXX to XXX)** externally. These models were superior to traditional models based on the American Society for Anesthesiologists (ASA) classification `r get_auc("ASA class")`, Charlson `r get_auc("Charlson")`, Elixhauser `r get_auc("Elixhauser")`, and the Rx Risk V `r get_auc("Rx Risk V")` comorbidity indices.

@@

We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration. The best models were then validated externally on a cohort from the Danish Hip Arthroplasty Register (DHR).

@@

## External validation cohort (Denmark)

@@

The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each.[@Quan2005] Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab. \@ref(tab:tabcategorization)). Some of the combined comorbidities were observed for less than 10 patients with or without PJI, and these diagnoses were not considered as potential predictors to reduce the risk of overfitting to spurious events. PJI was defined by the occurrence of relevant ICD-10 or procedural NOMESCO-codes recorded in the NPR within 90 days (Tab. \@ref(tab:infcodes)), or if the patient was reported to SHAR as re-operated due to infection within the same time frame.

@@

Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in a logistic LASSO regression model. We kept the $\lambda$ that minimized the mean cross-validated deviance. 100 bootstrap samples were then drawn from the observed data set.[@Austin2004]

Logistic LASSO regression was performed for each sample using the chosen $\lambda$ as penalty term. Regression coefficients were ranked based on their absolute values as a measure of variable importance. We then used Bayesian methods to estimate a posterior distribution and to identify two change points in a linear segmented piecewise regression model. The second change point was then used as a break-point. Influential predictors with absolute coefficients value above this point, scored a point. The whole process, each with 100 Bootstrap replicates, was repeated 100 times (yielding 10,000 bootstrap samples in total). Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables. Even more parsimonious models with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison. Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison. The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery. This data was retrieved from the Swedish national prescription register and classified according to Pratt et al.[@Pratt2018] Odds ratios for the final models were estimated with 95 % confidence intervals.

@@

Each developed model was used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation). Receiver operating characteristic (ROC) curves and their corresponding areas under those curve (AUC) were used to describe discrimination with 95 % confidence intervals that were based on percentiles from 2,000 non-parametric bootstrap samples. The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed. Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] The derived models were externally validated on a Danish cohort, and re-calibration of model intercepts was performed to account for different PJI incidences in Sweden compared to Denmark.

@@

We built an online web calculator available at **XXX** to be used in clinical practice. `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.

@@

We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). `r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of the patients developed a PJI within 90 days (Tab. \@ref(tab:tab1)). The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V. In addition, **XXX** patients from **XXX** were included as an external validation cohort (Fig. \@ref(fig:flowchart), right panel).

@@

Patients with `r excl_factors` were uncommon, wherefore those comorbidities were excluded as potential predictors prior to further modeling. ROC-curves (Fig. \@ref(fig:rocs)) and AUC-values (Fig. \@ref(fig:aucci)) for the main and reduced models were very similar, and for simplicity, we therefore focused on the reduced models. Estimated probability density curves for patients with and without PJI were partially overlapping, although patients with PJI had on average higher predicted probabilities for this outcome (Fig. \@ref(fig:sep)).

The reduced model for the prediction of PJI within 90 days in the Swedish cohort included `r coefs_text` (Tab. \@ref(tab:coefs)). This reduced model for the predicition of PJI within 90 days had an `r get_auc("Reduced model", "2")`, whereas ASA class `r get_auc("ASA class")`, the Charlson comorbidity index `r get_auc("Charlson")`, the Elixhauser comorbidity index `r get_auc("Elixhauser")`, and Rx Risk V `r get_auc("Rx Risk V")` had less discriminative ability (Fig. \@ref(fig:rocs) and \@ref(fig:aucci)). The reduced model had good accuracy as visualized by the calibration plot (Fig. \@ref(fig:calibration), left panel).

@@

We found that a multivariable main effects logistic regression model based on some easily identifiable clinical conditions, gender, and BMI was considerably better at predicting PJI within 90 days after THA than models based on either ASA class or the Charlson, Elixhauser or Rx Risk V comorbidity indices. This model also performed better than the established comorbidity measures in terms of AUC and calibration.

@@

Since the prediction model was based on easily accessible information it can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab. \@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics. For example, `r examples$desc` for PJI within 90 days.

@@

The predictors in our models were chosen based on a strict analytical approach, not on proposed relevance of patient or procedure characteristics. The majority of the identified predictor variables are, however, associated with the risk of developing PJI in previously developed prediction models and within the setting of observational studies, which is supportive of our own models.

@@

NPH initiated the study. OR managed the ethical review board application in Sweden and ABP in Denmark. EB developed the statistical model. ITA and UH performed the external validation. NPH and EB drafted, and all authors edited and finalized the manuscript.

@@

```{r brlprop}
coefs_selected_tab %>% 
knitr::kable(
  caption = "Variables selected by the bootstrap ranking procedure. Variables selected at least 10 out of 100 times were used in the main model. Variables chosen at least 80 times were kept in the reduced model as well. (BMI = body mass index. ASA class = American Society for Anesthesiologists classification. CNS = central nervous system.)")
```

@@

```{r coefs}
tbl_coefs_present %>% 
  mutate(
    beta = round(beta, 2)
  ) %>% 
knitr::kable(
  caption = 'Estimated coefficients (beta) and odds ratios (OR) with 95 % confidence intervals for the reduced models. BMI = body mass index ("under/normal weight as baseline). CNS = central nervous system. Primary osteoarthritis was baseline for diagnosis and female was baseline for sex.'
)
```
